---
title: Trick or tips 001 {R}
author: [kevin]
date: 2017-06-30
tags: [R, tips, trickortips]
rpkgs: [base, utils, graphics]
tweet: First post of the Fish and Tips series
draft: false
output:
  rmarkdown::html_page:
    fig_width: 3
    dev: svg
---


`r htmltools::HTML("{{< trickortips >}}")`


```{r codeChunkSetUp, include = FALSE}
source('../codeChunkSetUp.R')
```


---

## *Fish and tips* 0001

For the first post of the series, let's introduce 5 **R** tips:

  1. The %in% operator
  2. The `paste0()` function
  3. Using "\\r" in `cat()`
  4. The `sprintf()` function
  5. The `xaxs` and `yaxs` arguments of `par()`



## The '%in%' operator

I have known this for a while, but I keep remembering how it eases my pain when I
learned how to use it. Let's consider two vector `vecA` and `vecB`, the
`vecA%in%vecB` returns a vector of logical of the length `vecA` stating whether
the elements of `vecA` are in `vecB`:

```{r inOperator}
vecA <- c("a", "d", "2", "4", "e", ";")
vecB <- letters # vector of lower case letters the Roman alphabet
vecA%in%vecB
```

the i^th^ element of the output vector is `TRUE` if the i^th^ element of `vecA` is
among the elements of `vecB`. Note that you can have a similar result calling the `intersect()` functions which returns elements that `vecA` and `vecB` are sharing.

```{r intersect}
intersect(vecA, vecB)
```

By the way, you should have a look to the documentation of `intersect()`.



## The `paste0()` function

I often concatenate strings. The natural way to do this is to use the `paste()` function. Let's say I need to create a set of original file names such as: "myfile1", "myfile2", "myfile3", etc. I can call `paste()` like so:

```{r paste1}
paste("myfile", 1:7)
```

BUT, there is a annoying space between "myfile" and the number, to remove the
space, `sep` argument is needed:

```{r paste2}
paste("myfile", 1:3, sep="")
```

UNLESS you use the `paste0()` that is the version of `paste()` with the desired
default parameters:

```{r paste0}
paste0("myfile", 1:3)
```

I know... this means you'll save 5 previous characters each time you call `paste0()`!




## Using argument '\\r' in `cat()`

*Note that for this tip, you should execute the code-chunk in your own R console*

When programming we often need to monitor what we are doing, especially for
time-consuming calculation. Using `cat()` with all its default is very helpful
to report information on progress:

```{r cat, eval=F}
for (i in 1:10) {
  cat("iteration n°", i, "\n")
}
```

Here `"\n"` is what breaks the line! Instead of breaking the line, we can erase
the line, i.e. using `"\r"`:


```{r catErase, eval=F}
for (i in 1:10) {
  Sys.sleep(.25) # I added a pause so you can see what's happening
  cat("iteration n°", i, "\r")
}
```

I hope this gives you some ideas :). If you are thinking "where the *!&)@(* is
he going with this!" let me give you two extra way to monitor a for loop:

1. Use percentage:

```{r catMonitorPerc, eval=F}
nmx <- 222
for (i in 1:nmx) {
  Sys.sleep(.01) # replace it by what you are doing
  cat("==>", format(100*i/nmx, digits=4), "%", "\r")
}
```

2. Create you own progress bar:

```{r catMonitorPB, eval=F}
nmx <- 222
sz <- 60
for (i in 1:nmx) {
  Sys.sleep(.01) # replace it by what you are doing
  nb <- floor(sz*i/nmx)
  pgr <- paste(rep(c("="," "), c(nb, sz-nb)), collapse="")
  cat("|", pgr, "| ", "\r")
}
```

I hope you like this! Otherwise, stick to use `getTxtProgressBar()` and `setProgressBar()` functions, but I swear you are missing something.



## The `sprintf()` function

The function is a wrapper for a well-known C function `sprintf()`. There are
so much to say about it and I won't (so please have a careful look at the
documentation) but I think it is worth a tip to present the most used features
of this function.

The first argument of `sprintf()` is a string that includes tag to be replaced
by the extra arguments. A tag starts with a `%`, is followed by a
couple of optional characters tuning the display and ends with a letter
specifying the nature of the variable(s) to be displayed (number or string). The
most common letters are:

- "d" for integer value;
- "f" for double precision value;
- "s" for strings.

The most common options are:

- "n", an integer values specifying the width of the field;
- "0" indicates whether zero-padding should be used;
- ".m" where n is an integer.


That being said, to be as clear as possible, let me exemplify!

1. Different kind of variables:

```{r sprintf1, results='hold'}
sprintf("%d", 1)
sprintf("%f", pi)
sprintf("%s", "A")
```

2. Combine different variables:

```{r sprintf2}
sprintf("An integer: %d, a double: %f, a string: %s", 1, pi, "A")
```

3. Tune the field width:

```{r sprintf3}
sprintf("%10f", pi)
```

4. Zero-padding:

```{r sprintf4}
sprintf("%010f", pi)
```

5. Tune the precision:

```{r sprintf5}
sprintf("%.50f", pi)
```

Pretty cool, isn't it? To conclude two questions. Question 1: do you remember the file naming example using to showcase `paste0()`? I assume your answer if yes. Let's say you now have more than 100 files to name and you want to use zero-padding and a constant field width of 3 (e.g. "myfile001", "myfile010", "myfile100"). This leads
me to question 2: what do you do?  => Go `sprintf()`!

```{r sprintf6}
paste0("myfile", sprintf("%03d", c(1,10,100)))
```



## The `xaxs` and `yaxs` arguments of `par()`

This is also something I learnt years ago and keep using for almost all my
plots ever since. Default values of graphical parameters generate plot that
add and extra 4% to the axes limits. This is actually useful to ensure that
all graphical symbols are entirely displayed. However, I prefer using round
values as axes limits and handling myself their values so that all symbols stay visible. In order to get rid of the extra 4% added on the x-axis (y-axis), you
have to pass `xaxs="i"` (`yaxs="i"`) to the `par()` function. I compare below
a plot using the default parameters' values and the same plot using the tip:

```{r xaxsYaxs, fig.width=9.5}
par(mfrow=c(1,2), bty="l")
plot(0:10, 0:10, pch=20)
par(xaxs="i", yaxs="i")
plot(0:10, 0:10, pch=20)
```

Can you tell the difference? I now add two extra lines to stress the differences out.

```{r xaxsYaxsV2, fig.width=9.5}
par(mfrow=c(1,2), bty="l")
plot(0:10, 0:10, pch=20)
abline(v=c(0,10), h=c(0,10), col=2)
par(xaxs="i", yaxs="i")
plot(0:10, 0:10, pch=20)
abline(v=c(0,10), h=c(0,10), col=2)
```

**That's all folks... until next time!**

<details>
<summary> Session info <i class="fa fa-cogs" aria-hidden="true"></i> </summary>
```{r sessionInfo}
sessionInfo()
```
</details>
