---
title: R in Space - Geometry manipulation
author: [david, marieh, nicolas, kevin, elliot, steve]
date: 2018-04-02
tags: [R in Space, R, Spatial]
rpkgs: [sp, sf, raster]
draft: true
tweet: "R in Space - Spatial Objects in R"
estime: 10
navposts:
  prev: rinspace_spobjinr
  home: rinspace_homepage
  next: rinspace_rdataresources
output:
  rmarkdown::html_page:
    fig_width: 3
    dev: svg
---

![](https://img.shields.io/badge/inSileco-InDevelopment-3fb3b2.svg)

```{r codeChunkSetUp, echo=FALSE}
source('../../../static/Rscript/codeChunkSetUp.R')
```



To manipulate geometries, there are few options. For spatial objects defined
with `sp`, there is a specific package to manipulate vector objects: `rgeos`.
The equivalent of most common functions in `sp` are found in `sf`.
Finally, there are a set of interesting functions in `raster` to handle
the manipulation of rasters with vector objects.  


Let's first load the packages required and download a simple shapefile of Belgium.

```{r, message = F}
library(sp)
library(sf)
library(rgeos)
library(raster)
bel2 <- getData(name="GADM", country="BEL", path="data", level = 2)
plot(bel2)
```


## `sp` objects



### Subset

A quick subset:

```{r, message = F}
## Create a subset (A sub-shapefile)
slc <- c(8,11,12)
bel_slc <- bel2[slc,]
class(bel_slc)
```

Let's plot our selection:

```{r, message = F}
plot(bel_slc)
```


### Unions

Let's do the union of a selection of spatial polygons:

```{r, message = F}
bel_south <- gUnionCascaded(bel_slc)
bel_north <- gUnionCascaded(bel2[-slc,])
bel_one <- gUnionCascaded(bel2)
par(mfrow=c(1,3), mar=c(1,1,1,1))
plot(bel_south)
plot(bel_north)
plot(bel_one)
```

Let's combine them on one plot:

```{r, message = F}
plot(bel2, lwd = .1)
plot(bel_south, add=T, col=2)
plot(bel_north, add=T, col=4)
plot(bel2, lwd = .5, add = T)
plot(bel_one, lwd = 2, add = T, border = "grey55")
```




### Buffers

Buffer must be done on planar coordinates.

```{r, message = F}
plot(bel2)
bufs <- gBuffer(bel_south, width=0.4)
bufn <- gBuffer(bel_north, width=0.1)
plot(bufs, add=T, lwd=2, lty=2)
plot(bufn, add=T, lwd=3)
```

### Intersections

Intersections between `bel_south` and `bel_north`:

```{r, message = F}
par(mfrow = c(1,2))
plot(bel_one)
plot(gIntersection(bel_south, bel_north), col=5, add=T, lwd=2)
plot(bel_one)
plot(gIntersection(bufs, bufn), col=5, add=T, lwd=2)
```



### Differences

```{r}
bel_diff <- gDifference(bufs, bufn)
plot(bel_one)
plot(bel_diff, add=T, lwd=2, lty=3, col=2)
plot(bel_south, add=T, col=2)
```


## Overlays

One very useful tool is the **over()** function which provides a consistent
spatial overlay. For instance, here, I create a set of random points and
look which one are in `bel2`, `bel_north` and `bel_south`:

```{r}
pts <- SpatialPoints(
  coords = cbind(
    runif(50, bel2@bbox[1, 1], bel2@bbox[1, 2]),
    runif(50, bel2@bbox[2, 1], bel2@bbox[2, 2])
  ),
  proj4string = bel2@proj4string
)
# Make a SpatialPolygons
bel2_geom <- SpatialPolygons(bel2@polygons, proj4string = bel2@proj4string)
# plots
plot(bel2_geom)
points(pts, col = is.na(over(pts, bel2_geom))+2, pch =19)
```


## `sf` objects

Have a look at the dedicated vignette https://cran.r-project.org/web/packages/sf/vignettes/sf3.html


## `raster` objects

Let's now download a simple raster of eleveation.


```{r, message = F}
bel_elv <- getData(name="alt", country="BEL", path="data", level = 2)
plot(bel_elv)
```


### Extract


### Mask
