---
title: "Make your plot Hylian"
author:
  - Kevin Cazelles,
    David Beauchesne &
    Nicolas Casajus
status: finalVersion
pgmlang: R(>=3.0.0)
download:
  - text: "fontIcons"
    href: "Triforce.ttf.zip"
    icon: "fa fa-file-archive-o"
keywords:
  - plot
  - R
  - icon
  - zelda
date: "April 14, 2017 -- last update on July 16, 2017 ()"
tweet: "Make your plot Hylian"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    include:
        after_body: [footer.html, disqus.html]
---



```{r addStatus, echo=FALSE, results="asis", message=FALSE}
  source("Rscripts/addStatus.R")
  addStatus("HylianStyle.Rmd")
```


<br/>

#### Below are presented the 6 steps you need to make your plot a Hylian one! **May the triforce be with you**!


<br/>

### 0. R Version

```{r, echo = TRUE, eval = TRUE}
sessionInfo()[[1]]$version.string
```

<br/>

### 1. Open R and download the font

The [*Triforce* font](https://zeldauniverse.net/media/dir) is available on line.
Alternatively, you may simply download and install it, if you proceed so,
skip step 1.

```{r, echo = TRUE, eval = FALSE}
# Create a directory < fonts >
dir <- "_tmp/dir"
dir.create("_tmp/fonts", showWarnings = FALSE)

# Download Triforce font < Triforce.ttf >
url <- "http://zeldauniverse.net/wp-content/uploads/2014/01/Triforce.ttf_.zip"
download.file(url, paste0("dir", basename(url)))
dir("dir")

# Unzip the file
untar(paste0("dir", basename(url)), exdir = "dir", compressed = "gzip")
dir("dir")

# Remove .zip
file.remove(paste0("dir", basename(url)))
dir("dir")
```

<br/>

### 2. Install the `showtext` package

```{r, echo = TRUE, eval = FALSE}
install.package("showtext", dependencies = TRUE)
library(showtext)
```

```{r, echo = FALSE, eval = TRUE}
library(showtext)
```

<br/>

###  3. Look at the list of fonts available locally

```{r, echo = TRUE, eval = TRUE}
showtext.auto()
font.files()
```

If you have installed the font on your computer, it should be in the list,
then you can skip step 4.

<br/>

### 4. Add a path to your font

If you have downloaded the font using R, you must notify showtext about to
folder `dir` where the font has been stored. To so do, use the `font.paths()`
function:

```{r, echo = TRUE, eval = FALSE}
font.paths("dir")
```

```{r, echo = FALSE}
font.paths("assets/HylianStyle")
```

<br/>

### 5. Add the *Triforce* font to your session:


Either way, the Triforce font should now be available

```{r, echo = TRUE, eval = TRUE}
# Is the TTF present?
font.files()[grep("Triforce", font.files())]
```

Let's add it to our session:

```{r, echo = TRUE, eval = TRUE}
font.add(family = 'Triforce', regular = 'Triforce.ttf')
font.families()
```

<br/>

### 6. Finally, let's make our Hylian plot!

```{r, echo = TRUE, eval = TRUE, fig.width = 9.5, fig.height = 7, fig.showtext = TRUE}
par(family = "Triforce", las = 1, cex.lab = 2, cex.main = 2, mar = c(4, 5, 5, 0))
plot(x = c(0, 1), y = c(0, 1), type = "n", xlab = "X-axis", ylab = "Y-axis")
title(main = "Hylian Style - Link must be proud of us!")
points(x = runif(7), y = runif(7), pch = "*", cex = 2)
points(x = runif(7), y = runif(7), pch = "$", cex = 2, col = "#347AB7")
points(x = 0.1, y = 0.1, pch = "%", cex = 6, col = "#347AB7")
```
