---
title: "Online ecology - Individual species description"
author: [david, kevin, remi]
date: 2017-06-19
tags: [R, ecology, open-data, open-science]
rpkgs: [biomod2, kableExtra, knitr, magrittr, raster, rfishbase, rglobis, robis, sdmpredictors, sf, taxize]
draft: false
tweet: "Online ecology - Individual species description"
output:
  rmarkdown::html_page:
    toc: true
    fig_width: 3
    dev: svg
---


```{r codeChunkSetUp, include = FALSE}
source('../../../static/Rscript/codeChunkSetUp.R')
library(raster)
library(magrittr)
library(knitr)
library(kableExtra)
```


<br/>

# Online ecology

Let's imagine that we are interested in a species in a given area and wish to
know as much as possible about it. But, you can't go out in the field because
funding is running short. What we do have, however, is a certain knowledge of
the open data science tools that are at our disposal. In a series of post
about online ecology, we will find out exactly just how far these tools allow
us to delve into the ecology of the species that interest us.

<br/>

Online ecology posts:

<!-- 1. Individual species description (*this post*)
2. [Communities: cooccurrence](./docs/OnlineEcology-CommCoocc.html) (*under development*)
3. [Communities: interactions](./docs/OnlineEcology-CommInter.html) (*under development*)
4. Scenarios of global change impact on community structure (*under development*)
5. more to come! -->

<br/>

# Individual species description

This post focuses on the simplest yet still complicated aspect of ecology,
*i.e.* describing a species as thoroughly as possible with the tools at
our disposal.

Special thanks to the developers at [ROpenSci](https://ropensci.org/), who built
many of the R package used to access the open access tools we present in this post.

<br/>

## Setting up R

#### R version used to build the last update of this post

```{r infosession}
sessionInfo()[[1]]$version.string
```

<br/>

## Defining species and area of interest

We start by selecting the species and the area in which we are interested. For
this post, we focus on the Atlantic cod (*Gadus morhua*) in the estuary and
gulf of St. Lawrence in eastern Canada.

<br/>

Let's set the parameters and create the spatial bounding box that we will be
using for the area of interest. *Note that all these parameters can be changed
to extract information for other species in other habitats.*

<br/>

```{r parameters}
# Species of interest
    sp <- 'Gadus morhua'

# Extent of area of interest
    latmax <- 52.01312
    latmin <- 45.52399
    lonmax <- -55.73636
    lonmin <- -71.06333

# Create a spatial bounding box for the area of interest using the 'sf' package:
# create a matrix:
    bb <- cbind(c(lonmin,lonmax,lonmax,lonmin,lonmin),
                   c(latmin,latmin,latmax,latmax,latmin)) %>%
        # put that matrix into a list, because that's what `st_polygon()` needs
        list() %>%
        # Make the matrix a 'simple features' polygon:
        sf::st_polygon() %>%
        # and let's make it a simple feature column and give it information about the projection:
        sf::st_sfc(crs="+proj=longlat +datum=WGS84") %>%
        # finally, let's put the sfc in a simple features data.frame in the variable `geometry`:
        sf::st_sf(name="Study Site",geometry=.)
```

<br/>

## Describe your species

### Retrieve miscellaneous ecological information: ***Fishbase***

We'll start with a description of the species. First, let's see what [fishbase](http://www.fishbase.ca/) has to offer. This online data repository,
along with [sealifebase](http://www.sealifebase.org/), contains a lot of precious
information on marine and aquatic species all over the world and is accessible
through the package [`rfishbase`](https://github.com/ropensci/rfishbase)

<br/>

```{r ecology, eval = TRUE, results = 'asis'}
# Species ecology
ecol <- rfishbase::ecology(sp)
ecol <- cbind(colnames(ecol), t(ecol))
rownames(ecol) <- NULL
ecol <- ecol[ecol[,2] != 0, ] # remove 0
ecol <- ecol[!is.na(ecol[,2]), ] # remove NAs

knitr::kable(ecol, col.names = c('Descriptors', 'Attributes'), "html") %>%
    kableExtra::kable_styling(full_width = FALSE)
```


<br/>

### Retrieving taxonomic information: ***Taxize***

We can also extract taxonomic information using the package [`taxize`](https://ropensci.org/tutorials/taxize_tutorial.html).
This package allows you to extract and validate, among other things, the
taxonomy of millions of species by accessing an important number of online
databases accessible through their Application Programming Interface (API).

<br/>

```{r taxonomy1, results='asis'}
# Export the taxonomy of the species of interest
taxize::classification(sp, db = 'worms', accepted = TRUE, verbose = FALSE)[[1]] %>%
    knitr::kable("html") %>% kableExtra::kable_styling(full_width = FALSE)
```
<br>

```{r taxonomy2, eval=T}
# Retrieve a tsn for Gadus morhua, i.e. an unique identifier from the itis db
idtsn <- taxize::get_tsn(sp, accepted=TRUE, verbose = FALSE, row=1)[1]
```


```{r taxonomy3, eval=T}
# We can also extract the common or scientific names using sci2comm() & comm2sci(), respectively.
taxize::sci2comm(taxize::as.tsn(idtsn), db = 'itis')

# Or find out whether there are other names under which the species is known
taxize::synonyms(taxize::as.tsn(idtsn), db = 'itis')
```

```{r taxonomy4, results='as.is', eval=T}
# Another really interesting feature is to extract all known species at a given
# taxonomic scale. With the itis db, you should first find the tsn associated
# with Gadus. Using `taxize::get_tsn('gadus')` you'll find out that it is 164710
knitr::kable(taxize::children(164710, db = 'itis')[[1]], "html") %>%
  kableExtra::kable_styling(full_width = FALSE)
```

Sounds like this is consistent with [what Wikipedia says](https://en.wikipedia.org/wiki/Gadus)!



<br/>

### Retrieving trophic informations: ***GloBI***

We can also retrieve information on known biotic interactions involving our species
of interest. The [Global Biotic Interactions](http://globalbioticinteractions.org)
web platform contains thousands of empirical binary interactions for multiple types
of interactions, all over the world, and is accessible using the package [`rglobi`](https://github.com/ropensci/rglobi).

<br/>

```{r trophic1, results='asis'}
# There are multiple types of interactions available on GloBI
    knitr::kable(rglobi::get_interaction_types()[,1:3], 'html') %>%
        kable_styling(full_width = FALSE)
```

```{r trophic2}
# For now let's focus on predator-prey interactions
    prey <- rglobi::get_prey_of(sp)$target_taxon_name # Retrieve prey
    pred <- rglobi::get_predators_of(sp)$target_taxon_name # Retrieve predators
    length(prey) # Number of prey
    length(pred) # Number of predators
    prey[1:20] # First 20 prey
    pred[1:20] # First 20 predators
```

<br/>

## Making our search spatially explicit

Since we are interested in a specific area, making our search spatially explicit
would be highly valuable. Luckily, there are tools that allow us to do just that.

<br/>

### Retrieving: occurrence data: ***OBIS*** & ***GBIF***

OBIS is the [Ocean Biogeographic Information System](http://www.iobis.org/)
and their vision is: "To be the most comprehensive gateway to the world’s ocean
biodiversity and biogeographic data and information required to address pressing
coastal and world ocean concerns." We can get access to their **HUGE** database
through the [`robis`](https://github.com/iobis/robis) package.

Similarly, GBIF is the [Global Biodiversity Information Facility](http://www.gbif.org/)
and it aims at providing free and open access to biodiversity data. This open
source platform can be accessed through the [`rgbif`](https://github.com/ropensci/rgbif)
package.

We only cover the OBIS package in this post since the targeted species is marine,
but visit [`rgbif`](https://github.com/ropensci/rgbif) github repository for more
information on its use.

<br/>

```{r, OBIS data, eval=TRUE, fig.width=9.5}
# Download occurrence data for species and area of interest between 2010 and 2017
    OBIS <- robis::occurrence(scientificname = sp, geometry=sf::st_as_text(bb$geometry), startdate = as.Date("2010-01-01"), enddate = as.Date("2017-01-01"))

# Remove duplicates
    OBIS <- unique(OBIS)

# Transform as spatial file
    OBIS <- sf::st_as_sf(OBIS,
                     coords = c("decimalLongitude", "decimalLatitude"),
                     crs="+proj=longlat +datum=WGS84",
                     remove=FALSE)

# Visualize with mapview
    mapview::mapview(OBIS, cex = 4)@map
```


<br/>

## Retrieving environmental data: ***sdmpredictors***

There are multiple resources available to access environmental data directly
through R. Bathymetry data can be accessed using the
[`marmap`](https://github.com/ericpante/marmap) package. Other environmental
data can also be accessed using other packages like
[`raster`](https://cran.r-project.org/web/packages/raster/index.html)
for terrestrial climatic data (*i.e.* bioclim) and
[`rnoaa`](https://github.com/ropensci/rnoaa) for some environmental
layers like sea surface temperature and sea ice cover. Finally, a precious
resource is the recently published [`sdmpredictors`](https://cran.r-project.org/web/packages/sdmpredictors/index.html)
package, which provides access to multiple important global environmental datasets for marine
and terrestrial environments and a total of
`r nrow(sdmpredictors::list_layers(sdmpredictors::list_datasets()))`

<br/>

```{r, sdmpredictors}
# List sdmpredictors datasets
    knitr::kable(sdmpredictors::list_datasets()[,1:4], 'html') %>%
        kable_styling(full_width = FALSE)
  # issues with the above if all columns included
```

<br/>

We can then import environmental data using the `sdmpredictors()` function.
Since our study area of interest is marine, we will select environmental
variables accordingly from the `Bio-ORACLE` dataset, although data from `MARSPEC`
could also be used.

<br/>


```{r importEnvVar}
# Import marine layers
  layers <- c('BO_chlomean','BO_dissox','BO_ph','BO_salinity','BO_sstmean','BO_bathymean')
# layers <- 'BO_ph'
  dir.create("downloaded_predictors/", showWarnings = FALSE)
  envCov <- sdmpredictors::load_layers(layers, datadir = "downloaded_predictors/", rasterstack = TRUE)
```


<br/>

Clip them and plot them:

<br/>

```{r cropEnvCov, eval = TRUE}
# Clip to study area extent
envCov <- raster::stack(
    raster::crop(envCov, raster::extent(lonmin, lonmax, latmin, latmax))
  )
```

```{r cropEnvCov2, eval = TRUE}
# Data description
knitr::kable(sdmpredictors::get_layers_info(layers)$current[,1:4], 'html') %>%
  kable_styling()
```

```{r, plotEnvVar, eval = TRUE, fig.width = 9.5, fig.height = 6}
# Plot 'em!
  raster::plot(envCov)
```



<br/>

## Evaluating species distribution models: ***biomod2***

Now that we have species occurrence and environmental data, we can evaluate
species distribution models (SDMs). There are multiple ways and
packages to perform SDMs and this post [vignette](https://cran.r-project.org/web/packages/dismo/vignettes/sdm.pdf)
provides an extensive overview of the different methods available. For this
blog post, we will be using the package [`biomod2`](https://cran.r-project.org/web/packages/biomod2/index.html),
which implements most of the methods presented in the vignette. There is also
a thorough explanation of our to use `biomod2` in another very interesting [vignette](https://r-forge.r-project.org/scm/viewvc.php/*checkout*/pkg/biomod2/inst/doc/Simple_species_modelling.pdf?root=biomod).


<br/>

### Formatting the data

The first thing to do is to format our data to be able to work with `biomod2`
functions, which work with `BIOMOD.formated.data.PA` class objects. This entails
stacking all of our environmental data together to perform analyses. Luckily
this was accomplished previously when extracting environmental data using the
`sdmpredictors` package. You can then create your `biomod2` data class...

<br/>

```{r, SDM data, eval = TRUE, results = 'hide'}
SDMdata <- biomod2::BIOMOD_FormatingData(resp.var = rep(1,nrow(OBIS)),
                                        expl.var = envCov,
                                        resp.xy = data.frame(OBIS$decimalLongitude, OBIS$decimalLatitude),
                                        resp.name = sp,
                                        PA.nb.rep = 1)
```

<br/>

...use this object with the `biomod2` functions to generate a distribution
model for your species of interest...

<br/>

```{r SDM_model, eval = TRUE, results = 'hide', warning = FALSE}

SDM <- biomod2::BIOMOD_Modeling(
            data = SDMdata,
            models = "RF",
            # models = "MAXENT.Tsuruoka",
            # Basic options for modelling
            model.options = biomod2::BIOMOD_ModelingOptions(),
            SaveObj = FALSE)
```


<br/>

...extract the evaluation of the fit of your model...

<br/>

```{r, modelEval}
# Model evaluation
  knitr::kable(biomod2::get_evaluations(SDM), 'html')  %>%
      kable_styling(full_width = FALSE)
```


<br/>

... and finally, once the models are calibrated and evaluated, you can project your
species spatial distribution:

<br/>

```{r, sdmSpatial, results = 'hide', fig.width = 7, fig.height = 7}
# Model spatial projection
  SDMproj <- biomod2::BIOMOD_Projection(modeling.output = SDM,
                                        new.env = envCov,
                                        proj.name = 'Awesome!',
                                        selected.models = 'all',
                                        binary.meth = 'TSS',
                                        compress = 'xz',
                                        clamping.mask = F,
                                        output.format = '.grd')
# ... and visualize them!
  biomod2::plot(SDMproj)
```

<!-- unlink('./Gadus.morhua/', recursive = T) -->



<br/>

Riveting!



<br/>

## Concluding remarks

Starting with a very basic idea, *i.e.* describing the ecology of a species in
a given area, we ended up with a species description, some known occurrences,
environmental data describing our area of interest and the potential spatial
distribution of our species!

...but should we stop there?

The reality is, there is so much more going on in our area of interest than
the spatial distribution of a single species. In the [next post](OnlineEcology-Communities.html),
we will look into community scale analyses using some of the tools presented in
this post, alongside some new ones specifically designed for community ecology.



<details>
<summary> Session info <i class="fa fa-cogs" aria-hidden="true"></i> </summary>
```{r sessionInfo}
sessionInfo()
```
</details>
