---
title: R in Space - Basic mapping
author: [david, marieh, nicolas, kevin, elliot, steve]
tags: [R in Space, R, Spatial]
draft: true
tweet: "R in Space - Basic mapping"
estime: 10
date: 2018-04-03
navposts:
  prev: rinspace_import
  home: rinspace_homepage
  next: rinspace_spobjinr
output:
  rmarkdown::html_page:
    fig_width: 3
    dev: svg
---

![](https://img.shields.io/badge/inSileco-InDevelopment-3fb3b2.svg)


## Plot methods for `sp` objects

Let's first get some data:

```{r getdata, message = F}
library(sp)
library(sf)
library(raster)
dir.create("data")
bel1 <- getData("GADM", country = "Belgium", level = 1, path = "data/")
bel2 <- getData("GADM", country = "Belgium", level = 2, path = "data/")
```

First plot:

```{r plotsp1}
plot(bel1)
```

First two plots:

```{r plotsp2}
par(mfrow = c(1,2), mar = c(0,0,0,0))
plot(bel1, col = 1:3)
plot(bel2, lty = 2)
```

First two plots:

```{r plotsp3}
plot(bel1, col = "grey85", main = "Belgium")
plot(bel2, lwd = .5, add = T)
points(coordinates(bel2), pch = 19, cex = .4)
city = coordinates(bel2)[1,]
text(city[1], city[2], labels = "Bruxelles", pos = 3, col = "#9728eb")
```




## plot methods for `sf` objects

Let's first get `sf` object:

```{r map_sf0, message = F}
bel1_sf <- st_as_sf(bel1)
bel2_sf <- st_as_sf(bel2)
```

Let's try a simple plot.

```{r map_sf1, message = F}
plot(bel1_sf)
```

It actually plot one map per variables and color the polygon according
to the values. To obtain something similar to what is found in `sp`,
an `sfc` object must be used:


```{r map_sf2, message = F}
plot(st_geometry(bel1_sf), lwd = 1.5, col = "grey85")
plot(st_geometry(bel2_sf), lwd = .5, add = T)
```




## plot methods for `raster` objects


## Using ggmap


## mapview

```{r mapview, message = F}
library(mapview)
mapview(bel1)
```
