---
title: "Fish and tips 01 {R}"
author: Kevin Cazelles
date: 2017-06-30
tags: ["R","tips"]
draft: false
output:
  rmarkdown::html_page:
    toc: true
    fig_width: 3
    dev: "svg"
---


## Code tips you said?

Ever tumbled on a code chunk that made you say: "I should have known this
*¶ø?!@~:&* piece of code long ago!"? Chances are you have, frustratingly,
just like we have, and on multiple occasions too.

In comes *blogPostSeriesName*!

*blogPostSeriesName* is a series of blog posts that each present 5 -- hopefully
helpful -- coding tips for a specific programming language.

Posts should be short (*i.e.* no more than 5 lines of code, max 80 characters
per line, except when appropriate) and provide tips of many kind: a function,
a way of combining of functions, a single argument, a note about the philosophy
of the language and practical consequences, tricks to improve the way you code,
good practices, etc.

Note that while some tips might be obvious for careful documentation readers
(God bless them for their wisdom), almost all the tips we present constituted
some sort of eureka moments for us.

By the way, there are undoubtedly similar initiatives on the web (*e.g.*
['One R Tip a Day' account](https://twitter.com/RLangTip)). Also, feel free to
[send us](mailto:kevin.cazelles@gmail.com) tip ideas or a post of code tips of your own which we will be happy to
incorporate to *blogPostSeriesName*!

Enjoy and get ready to frustratingly appreciate our tips!

---

## *blogPostSeriesName* 0001

For the first post of the series, let's introduce 5 **R** tips:

  1. The %in% operator      
  2. The `paste0()` function    
  3. Using "\\r" in `cat()`   
  4. The `sprintf()` function     
  5. The `xaxs` and `yaxs` arguments of `par()`     



## The '%in%' operator

I have known this for a while, but I keep remembering how it eases my pain when I
learned how to use it. Let's consider two vector `vecA` and `vecB`, the
`vecA%in%vecB` returns a vector of logical of the length `vecA` stating whether
the elements of `vecA` are in `vecB`:

```{r inOperator}
vecA <- c("a", "d", "2", "4", "e", ";")
vecB <- letters # vector of lower case letters the Roman alphabet
vecA%in%vecB
```

the i^th^ element of the output vector is `TRUE` if the i^th^ element of `vecA` is
among the elements of `vecB`. Note that you can have a similar result calling the `intersect()` functions which returns elements that `vecA` and `vecB` are sharing.

```{r intersect}
intersect(vecA, vecB)
```

By the way, you should have a look to the documentation of `intersect()`.



## The `paste0()` function

I often concatenate strings. The natural way to do this is to use the `paste()` function. Let's say I need to create a set of original file names such as: "myfile1", "myfile2", "myfile3", etc. I can call `paste()` like so:

```{r paste1}
paste("myfile", 1:7)
```

BUT, there is a annoying space between "myfile" and the number, to remove the
space, `sep` argument is needed:

```{r paste2}
paste("myfile", 1:3, sep="")
```

UNLESS you use the `paste0()` that is the version of `paste()` with the desired
default parameters:

```{r paste0}
paste0("myfile", 1:3)
```

I know... this means you'll save 5 previous characters each time you call `paste0()`!




## Using argument '\\r' in `cat()`

*Note that for this tip, you should execute the code-chunk in your own R console*

When programming we often need to monitor what we are doing, especially for  
time-consuming calculation. Using `cat()` with all its default is very helpful
to report information on progress:

```{r cat, eval=F}
for (i in 1:10) {
  cat("iteration n°", i, "\n")
}
```

Here `"\n"` is what breaks the line! Instead of breaking the line, we can erase
the line, i.e. using `"\r"`:


```{r catErase, eval=F}
for (i in 1:10) {
  Sys.sleep(.25) # I added a pause so you can see what's happening
  cat("iteration n°", i, "\r")
}
```

I hope this gives you some ideas :). If you are thinking "where the *!&)@(* is
he going with this!" let me give you two extra way to monitor a for loop:

1. Use percentage:

```{r catMonitorPerc, eval=F}
nmx <- 222
for (i in 1:nmx) {
  Sys.sleep(.01) # replace it by what you are doing
  cat("==>", format(100*i/nmx, digits=4), "%", "\r")
}
```

2. Create you own progress bar:

```{r catMonitorPB, eval=F}
nmx <- 222
sz <- 60
for (i in 1:nmx) {
  Sys.sleep(.01) # replace it by what you are doing
  nb <- floor(sz*i/nmx)
  pgr <- paste(rep(c("="," "), c(nb, sz-nb)), collapse="")
  cat("|", pgr, "| ", "\r")
}
```

I hope you like this! Otherwise, stick to use `getTxtProgressBar()` and `setProgressBar()` functions, but I swear you are missing something.



## The `sprintf()` function

The function is a wrapper for a well-known C function `sprintf()`. There are
so much to say about it and I won't (so please have a careful look at the
documentation) but I think it is worth a tip to present the most used features
of this function.

The first argument of `sprintf()` is a string that includes tag to be replaced
by the extra arguments. A tag starts with a `%`, is followed by a
couple of optional characters tuning the display and ends with a letter
specifying the nature of the variable(s) to be displayed (number or string). The
most common letters are:

- "d" for integer value;  
- "f" for double precision value;
- "s" for strings.  

The most common options are:

- "n", an integer values specifying the width of the field;
- "0" indicates whether zero-padding should be used;    
- ".m" where n is an integer.  


That being said, to be as clear as possible, let me exemplify!

1. Different kind of variables:

```{r sprintf1, results='hold'}
sprintf("%d", 1)
sprintf("%f", pi)
sprintf("%s", "A")
```

2. Combine different variables:

```{r sprintf2}
sprintf("An integer: %d, a double: %f, a string: %s", 1, pi, "A")
```

3. Tune the field width:

```{r sprintf3}
sprintf("%10f", pi)
```

4. Zero-padding:

```{r sprintf4}
sprintf("%010f", pi)
```

5. Tune the precision:   

```{r sprintf5}
sprintf("%.50f", pi)
```

Pretty cool, isn't it? To conclude two questions. Question 1: do you remember the file naming example using to showcase `paste0()`? I assume your answer if yes. Let's say you now have more than 100 files to name and you want to use zero-padding and a constant field width of 3 (e.g. "myfile001", "myfile010", "myfile100"). This leads
me to question 2: what do you do?  => Go `sprintf()`!

```{r sprintf6}
paste0("myfile", sprintf("%03d", c(1,10,100)))
```



## The `xaxs` and `yaxs` arguments of `par()`

This is also something I learnt years ago and keep using for almost all my
plots ever since. Default values of graphical parameters generate plot that
add and extra 4% to the axes limits. This is actually useful to ensure that
all graphical symbols are entirely displayed. However, I prefer using round
values as axes limits and handling myself their values so that all symbols stay visible. In order to get rid of the extra 4% added on the x-axis (y-axis), you
have to pass `xaxs="i"` (`yaxs="i"`) to the `par()` function. I compare below
a plot using the default parameters' values and the same plot using the tip:

```{r xaxsYaxs, fig.width=9.5}
par(mfrow=c(1,2), bty="l")
plot(0:10, 0:10, pch=20)     
par(xaxs="i", yaxs="i")   
plot(0:10, 0:10, pch=20)   
```

Can you the see the difference? Let me add two extra code lines to stress the differences out.

```{r xaxsYaxsV2, fig.width=9.5}
par(mfrow=c(1,2), bty="l")
plot(0:10, 0:10, pch=20)  
abline(v=c(0,10), h=c(0,10), col=2)   
par(xaxs="i", yaxs="i")   
plot(0:10, 0:10, pch=20)   
abline(v=c(0,10), h=c(0,10), col=2)
```

<br/>

**That's all folks, until next time!**

<br/>
