---
title: Gantt charts in R
author: [david]
date: 2017-09-20
tags: [R,time management,gantt]
draft: false
tweet: Gantt charts in R
output:
  rmarkdown::html_page:
    toc: true
    toc_float:
      collapsed: false
    fig_width: 3
    dev: svg
---

![](https://img.shields.io/badge/inSileco-UnderReview-ffdd55.svg)

<!--
setwd('/users/davidbeauchesne/dropbox/phd/insileco.github.io/content/post/')
rmarkdown::render('./ganttR.rmd', 'html_document')
-->

<br/>

# Gantt charts

Gantt charts are a very useful way to organize projects into milestones and tasks visually. They are also dead useful when comes the time to create a timeline for a project, whether it be for a research project, field work or grant applications. I recently wanted to create one, but was unsatisfied by what was available out there in terms of visuals (although there are neat R examples, see [here](https://www.r-bloggers.com/gantt-charts-in-r-using-plotly/), [here](https://stackoverflow.com/questions/3550341/gantt-charts-with-r) and [here](https://davetang.org/muse/2017/02/03/gantt-chart-using-r/)). So I came up with my own!

For those who might be interested only in generating charts without all the code, jump to the end, copy the `ganttR` function I created and Gantt away!

<br/>

# Initiate R

```{r init, message = F, warnings = F}
rm(list=ls())
# R version
sessionInfo()[[1]]$version.string

# Packages required
library(knitr)
library(tidyverse)
library(RColorBrewer)
library(kableExtra)
```

<br/>

# The code!

I wanted to come up with something simple and visually appealing, so I created a figure that simply required the following elements: milestone, tasks, start date, due data and status.

## Data

<br/>

```{r data}
# Let's organize a fake manuscript project to generate the chart.
Manuscript <- c('Research & readings','Data preparation','Analyses','Plan','Introduction','Methods','Results','Discussion','Conclusion','Preliminary version to co-authors')
startDate <- c('2017-09-01','2017-09-14','2017-09-28','2017-09-21','2017-09-28','2017-09-14','2017-10-28','2017-11-14','2017-11-28','2017-12-11')
dueDate <- c('2017-11-28','2017-09-28','2017-10-28','2017-09-28','2017-10-28','2017-10-28','2017-11-14','2017-11-28','2017-12-04','2017-12-11')
status <- c('I','C','I','C','I','I','I','I','I','I')
nTasks <- length(Manuscript)

# Create the data frame.
# In statuses, "I" and "C" would stand for "Incomplete" and "Complete", respectively
df <- data.frame(milestones = rep('Manuscript', nTasks),
                 tasks = Manuscript,
                 startDate = as.Date(startDate),
                 dueDate = as.Date(dueDate),
                 status = status,
                 stringsAsFactors = FALSE)

kable(df, "html") %>%
    kable_styling(full_width = F)
```

<br/>

We can now create our Gantt chart!

<br/>

## The chart

<br/>

```{r singleGantt, fig.width = 10, fig.height = 4}
# Let's first organize our graph in order of tasks startDate
df <- arrange(df, desc(startDate))

# We need a date range for which we wish to crete the graph.
# Let's select the duration of the milestone
dateRange <- c(min(df$startDate), max(df$dueDate))

# We also need the number of elements to add to our graph.
# It will be the number of tasks plus the number of milestones
nameMilestones <- unique(df$milestones)
nMilestones <- length(nameMilestones)
nLines <- nTasks + nMilestones

# We also need a date sequence that will be used as one of our axes
# We select the date range divided into 7 days periods
dateSeq <- seq.Date(dateRange[1], dateRange[2], by = 7)

# Finally, we need a color palette for the project
# We take three colors, the first will be for the milestones
# The second will be for incomplete tasks
# The third color will be for completed tasks
cols <- c('#4f739d','#4f739dBB','#4f739d33')

# Gantt chart
par(family = "serif", mar = c(6,9,2,0))
plot(x = 1, y = 1, col = 'transparent', xlim = dateRange, ylim = c(1,nLines), bty = "n",ann = FALSE, xaxt = "n",yaxt = "n",type = "n",bg = 'grey')

# Add axes, tasks and milestones
mtext(dateSeq, side = 1, at = dateSeq, las = 3, line = 1.5, cex = 0.75)
axis(1, dateSeq, labels = F, line = 0.5)
mtext(df$tasks, side = 2, at = 1:nrow(df), las = 1, line = 0, cex = 0.75)
mtext(nameMilestones, side = 2, at = nrow(df)+1, las = 1, line = 8, font = 2, adj = 0, cex = 0.8)

# Add tasks
for(i in 1:nTasks) {
    lines(c(i,i) ~ c(df$startDate[i], df$dueDate[i]), lwd = 6, col = if(df$status[i] == 'C') cols[3] else cols[2])
}

# Add milestone
lines(c(nLines,nLines) ~ c(min(df$startDate), max(df$dueDate)), lwd = 8, col = cols[1])

# Add today's date
abline(v = as.Date(format(Sys.time(), format = "%Y-%m-%d")), lwd = 2, lty = 2)
```

<br/>

## Multiple milestones

Now Gantt charts are also useful to visualize and organize multiple projects together, so let's make this chart multi-milestony.

<br/>

```{r multiGantt, fig.width = 10}
# We will simply duplicate the manuscript and do as though we are working on three consecutive manuscripts
df2 <- df3 <- df
df2$milestones <- 'Manuscript2'
df3$milestones <- 'Manuscript3'
df2[,c('startDate','dueDate')] <- df2[,c('startDate','dueDate')] + as.numeric(diff(dateRange))
df3[,c('startDate','dueDate')] <- df3[,c('startDate','dueDate')] + as.numeric(diff(dateRange))*2
df2$status <- df3$status <- rep('I',nTasks)
dfM <- rbind(df, df2, df3)

# ... and go through the process of setting up the elements needed for the chart
# Let's start with the colors this time
# We take three colors, the first will be for the milestones
# The second will be for incomplete tasks
# The third color will be for completed tasks
nameMilestones <- unique(dfM$milestones)
nMilestones <- length(nameMilestones)
cols <- data.frame(milestones = nameMilestones,
                   col = brewer.pal(nMilestones, "Dark2"),
                   stringsAsFactors = F)

# Let's organize our dataset to produce the graph
dfMulti <- dfM %>%
      group_by(milestones) %>% # group by milestones
      summarise(startDate = min(startDate),
                dueDate = max(dueDate)) %>% # Determine the beginning and end date of milestones
      mutate(tasks = milestones, status = 'M') %>% # Give a name and a status
      bind_rows(dfM) %>% # Bind milestones with tasks
      mutate(lwd = ifelse(milestones == tasks, 8, 6)) %>% # Add line width for graph
      left_join(cols, by = 'milestones') %>% # add colors
      mutate(col = ifelse(status == 'I', paste0(col,'BB'), col)) %>% # change colors according to status
      mutate(col = ifelse(status == 'C', paste0(col,'33'), col)) %>% # change colors according to status
      mutate(cex = ifelse(status == 'M', 0.8, 0.75)) %>%
      mutate(adj = ifelse(status == 'M', 0, 1)) %>%
      mutate(line = ifelse(status == 'M', 8, 0.5)) %>%
      mutate(font = ifelse(status == 'M', 2, 1)) %>%
      arrange(milestones,desc(startDate),dueDate) # sort table

# We need a date range for which we wish to crete the graph.
# Let's select the duration of the milestone
dateRange <- c(min(dfMulti$startDate), max(dfMulti$dueDate))

# We also need a date sequence that will be used as one of our axes
# We select the date range divided into 7 days periods
dateSeq <- seq.Date(dateRange[1], dateRange[2], by = 7)


# Gantt chart
nLines <- nrow(dfMulti)
par(family = "serif", mar = c(6,9,2,0))
plot(x = 1, y = 1, col = 'transparent', xlim = dateRange, ylim = c(1,nLines), bty = "n",ann = FALSE, xaxt = "n",yaxt = "n",type = "n",bg = 'grey')
mtext(dateSeq, side = 1, at = dateSeq, las = 3, line = 1.5, cex = 0.75)
axis(1, dateSeq, labels = F, line = 0.5)

for(i in 1:nLines) {
    lines(c(i,i) ~ c(dfMulti$startDate[i],dfMulti$dueDate[i]),
          lwd = dfMulti$lwd[i],
          col = dfMulti$col[i])
    mtext(dfMulti$tasks[i],
          side = 2,
          at = i,
          las = 1,
          adj = dfMulti$adj[i],
          line = dfMulti$line[i],
          cex = dfMulti$cex[i],
          font = dfMulti$font[i])
}

abline(h = which(dfMulti$status == 'M') + 0.5, col = '#634d42')
abline(v = as.Date(format(Sys.time(), format = "%Y-%m-%d")), lwd = 2, lty = 2)
```

<br/>

# ganttR()

Now that we have all the code, let's simply transform it into a reusable function. Let's also give it the choice to generate two types of graph: 1) all milestones and tasks or 2) only milestones.

```{r, ganttR, fig.width = 10, warning = F}
ganttR <- function(df, type = 'all') {

    nameMilestones <- unique(df$milestones)
    nMilestones <- length(nameMilestones)
    rbPal <- colorRampPalette(brewer.pal(8, "Dark2")) # Color palette
    cols <- data.frame(milestones = nameMilestones,
                       col = rbPal(nMilestones),
                       stringsAsFactors = F)
    cols <- cols[1:nMilestones, ]

    # Let's organize our dataset to produce the graph
    df <- df %>%
          group_by(milestones) %>% # group by milestones
          summarise(startDate = min(startDate),
                    dueDate = max(dueDate)) %>% # Determine the beginning and end date of milestones
          mutate(tasks = milestones, status = 'M') %>% # Give a name and a status
          bind_rows(df) %>% # Bind milestones with tasks
          mutate(lwd = ifelse(milestones == tasks, 8, 6)) %>% # Add line width for graph
          left_join(cols, by = 'milestones') %>% # add colors
          mutate(col = ifelse(status == 'I', paste0(col,'BB'), col)) %>% # change colors according to status
          mutate(col = ifelse(status == 'C', paste0(col,'33'), col)) %>% # change colors according to status
          mutate(cex = ifelse(status == 'M', 0.8, 0.75)) %>%
          mutate(adj = ifelse(status == 'M', 0, 1)) %>%
          mutate(line = ifelse(status == 'M', 8, 0.5)) %>%
          mutate(font = ifelse(status == 'M', 2, 1)) %>%
          arrange(milestones,desc(startDate),dueDate) # sort table

    # We need a date range for which we wish to crete the graph.
    # Let's select the duration of the milestone
    dateRange <- c(min(df$startDate), max(df$dueDate))

    # We also need a date sequence that will be used as one of our axes
    # We select the date range divided into 7 days periods
    dateSeq <- seq.Date(dateRange[1], dateRange[2], by = 7)


    # Gantt chart for 'all' type
    if(type == 'all') {
        nLines <- nrow(df)
        par(family = "serif", mar = c(6,9,2,0))
        plot(x = 1, y = 1, col = 'transparent', xlim = dateRange, ylim = c(1,nLines), bty = "n",ann = FALSE, xaxt = "n",yaxt = "n",type = "n",bg = 'grey')
        mtext(dateSeq, side = 1, at = dateSeq, las = 3, line = 1.5, cex = 0.75)
        axis(1, dateSeq, labels = F, line = 0.5)

        for(i in 1:nLines) {
            lines(c(i,i) ~ c(df$startDate[i],df$dueDate[i]),
                  lwd = df$lwd[i],
                  col = df$col[i])
            mtext(df$tasks[i],
                  side = 2,
                  at = i,
                  las = 1,
                  adj = df$adj[i],
                  line = df$line[i],
                  cex = df$cex[i],
                  font = df$font[i])
        }

        abline(h = which(df$status == 'M') + 0.5, col = '#634d42')
        abline(v = as.Date(format(Sys.time(), format = "%Y-%m-%d")), lwd = 2, lty = 2)
    }

    # Gantt chart for 'milestones' only
    if(type == 'milestones') {
        nLines <- nMilestones
        ms <- which(df$status == 'M')
        par(family = "serif", mar = c(6,9,2,0))
        plot(x = 1, y = 1, col = 'transparent', xlim = dateRange, ylim = c(1,nLines), bty = "n",ann = FALSE, xaxt = "n",yaxt = "n",type = "n",bg = 'grey')
        mtext(dateSeq, side = 1, at = dateSeq, las = 3, line = 1.5, cex = 0.75)
        axis(1, dateSeq, labels = F, line = 0.5)

        for(i in 1:nLines) {
            lines(c(i,i) ~ c(df$startDate[ms[i]],df$dueDate[ms[i]]),
                  lwd = df$lwd[ms[i]],
                  col = df$col[ms[i]])
            mtext(df$tasks[ms[i]],
                  side = 2,
                  at = i,
                  las = 1,
                  adj = 1,
                  line = 0.5,
                  cex = df$cex[ms[i]],
                  font = df$font[ms[i]])
        }
        abline(v = as.Date(format(Sys.time(), format = "%Y-%m-%d")), lwd = 2, lty = 2)
    }   
}

# Single milestone
ganttR(df)

# Multiple milestones
ganttR(dfM)

# Milestones only
ganttR(dfM, 'milestones')
```
