---
title: Make your plot Hylian
author: [david,kevin,nicolas]
date: 2017-04-14
tags: [plot,R,icon,zelda]
draft: false
tweet: Make your plot Hylian
output:
  rmarkdown::html_page:
    toc: true
    fig_width: 3
    dev: svg
---


#### Below are presented the 6 steps you need to make your plot a Hylian one! **May the triforce be with you**!


### 0. R Version

```{r, echo = TRUE, eval = TRUE}
sessionInfo()[[1L]]$version.string
```



### 1. Open R and download the font

The [*Triforce* font](https://zeldauniverse.net/media/fonts/) is available on line.
Below we download it using R! Alternatively, you may simply download and install
it using a couple of clicks, if you proceed so, skip step 1.


```{r, echo = TRUE, eval = TRUE}
# Create a directory < fonts >
dr <- "hylianStyle_files/"
dir.create(dr, showWarnings = FALSE)

# Download Triforce font < Triforce.ttf >
url <- "http://zeldauniverse.net/wp-content/uploads/2014/01/Triforce.ttf_.zip"
download.file(url, destfile=paste0(dr, basename(url)),method="libcurl")

# Unzip the file
unzip(paste0(dr, basename(url)), exdir = "./hylianStyle_files")

# Remove .zip
file.remove(paste0(dr, basename(url)))
```

### 2. Install the `showtext` package

```{r, echo = TRUE, eval = FALSE}
install.packages("showtext", dependencies = TRUE)
```

###  3. Look at the list of fonts available locally

```{r, echo = TRUE, eval = TRUE, R.options=list(max.print=30)}
library(showtext)
showtext_auto()
font_files()
```

If you have installed the font on your computer, it should be in the list,
then you can skip step 4.


### 4. Add a path to your font

If you have downloaded the font using R, you must notify showtext about to
folder `dir` where the font has been stored. To so do, use the `font_paths()`
function:

```{r, echo = TRUE, eval = TRUE}
font_paths(dr)
```

### 5. Add the *Triforce* font to your session:


Either way, the Triforce font should now be available

```{r, echo = TRUE, eval = TRUE}
# Is the TTF present?
font_files()[grep("Triforce", font_files())]
```

Let's add it to our session:

```{r, echo = TRUE, eval = TRUE}
font_add(family = 'Triforce', regular = 'Triforce.ttf')
font_families()
```

### 6. Finally, let's make our Hylian plot!

```{r, echo = TRUE, eval = TRUE, fig.width = 9.5, fig.height = 7, fig.showtext = TRUE}
par(family = "Triforce", las = 1, cex.lab = 2, cex.main = 2, mar = c(4, 5, 5, 0))
plot(x = c(0, 1), y = c(0, 1), type = "n", xlab = "X-axis", ylab = "Y-axis")
title(main = "Hylian Style - Link must be proud of us!")
points(x = runif(7), y = runif(7), pch = "*", cex = 4)
points(x = runif(7), y = runif(7), pch = "$", cex = 4, col = "#347AB7")
points(x = 0.1, y = 0.1, pch = "%", cex = 10, col = "#347AB7")
```
