---
title: Hierarchical edge bundling using base plot
author: [david]
reviewer: [kevin]
date: 2018-02-21
tags: [R, Viz, network]
package: [graphicsutils, dplyr, plotrix]
draft: false
tweet: "Hierarchical edge bundling using base plot"
estime: 10
---



So, in my thesis I work extensively with networks and coming up with meaningful and
useful visualization can be a bit of a challenge. I've recently begun toying with
the notion of using hierarchical edge bundling to represent the complexity of the
networks I work with and the different types of networks I am attempting to link
together. While this type of visualization can be close to useless at a fine scale
to study individual connections, it can be pretty useful to glean insights as to
the overall arrangements of links within networks.

With that in mind, I began going through the material available to make hierarchical
edge bundling in R. There is quite a bit of material already available on R, but
what I found was always through `ggraph` and `ggplot`, which I personally don't
enjoy using all that much (*e.g.*
[this](https://www.r-graph-gallery.com/hierarchical-edge-bundling/)).
I rather prefer having full control and understanding of what I am doing with my
visualizations. So, is here the product of this work to generate the figure I wished to
create for my thesis!

## Libraries

```{r param}
library(graphicsutils)
data(gpu)
```

## Data

Let's start by simulating data for a metanetwork (*i.e.* a network composed of
multiple networks). To make this as simple as possible, I will simply simulate
a dataset composed of hierarchical nodes and another one identifying the links
between those nodes.

```{r data}
# Individual network nodes
  opt <- options()
  options(stringsAsFactors = FALSE)
  drivers <- data.frame(network = 'Drivers', name = paste0('driver_',1:15))
  species <- data.frame(network = 'Species', name = paste0('species_',1:30))
  actions <- data.frame(network = 'Actions', name = paste0('action_',1:9))
  managers <- data.frame(network = 'Managers', name = paste0('manager_',1:9))
  beneficiaries <- data.frame(network = 'Beneficiaries', name = paste0('beneficiary_',1:9))

# Nodes dataframe
  nodes <- rbind(drivers, species, actions, managers, beneficiaries)

# Simulate links
  links <- data.frame(from = sample(nodes$name, 300, replace = T),
                      to = sample(nodes$name, 300, replace = T))
  links <- links[!duplicated(links), ]   # Remove duplicates
  links <- links[!links$from == links$to, ] # Remove "cannibalism"

# Combine in a single object
  metanetwork <- vector('list', 0)
  metanetwork$nodes <- nodes
  metanetwork$links <- links

head(nodes)
head(links)
```

## Circle functions

Now that we have the nodes and links, we need to create the necessary functions
that will allow us to position our data on a circular graph.

<br/>

### Circle coordinates

The first function we need will be used to generate the x and y coordinates of a
circle for a given angle and circle radius, which we can obtain with some circle trigonometry.

```{r coordCircle}
# Let's begin by creating a function that will give us the x and y coordinates
# of the outside of a circle given a certain radius
coordCircle <- function(theta = NULL, radius = 1) {
  data.frame(x = radius * cos(theta),
             y = radius * sin(theta))
  }

coordCircle(theta = pi, radius = 1)
```


<br/>

### Group boundaries

The second function that we need is one that will allow us to distribute groups
on the circular plot according to the number of elements composing that group.
In essence, we wish to provide the upper and lower boundaries of each individual
network composing the metanetwork. We may also want to add a gap between the
individual networks on the graph. This will make the visual aspect of the graph
more appealing.

```{r groupBoundaries}
bound <- function(metanetwork, gap = .025, addGap = T) {
  # Metanetwork list composed of "nodes" and "links"
  # Size of gap between groups on the graph
  # addGap logical whether to add gap or not  
  nGroup <- as.data.frame(table(metanetwork$nodes$network))
  nGroup$Prop <- nGroup$Freq / sum(nGroup$Freq)
  nGroup$spanDeg <- 2 * pi * nGroup$Prop
  nGroup$upper <- nGroup$lower <- 0
  for(i in 2:nrow(nGroup)) nGroup$lower[i] <- nGroup$lower[i-1] + nGroup$spanDeg[i-1]
  nGroup$upper <- nGroup$lower + nGroup$spanDeg

  if (addGap) {
    nGroup$lower <- nGroup$lower + gap/2
    nGroup$upper <- nGroup$upper - gap/2
  }

  nGroup
}

# Add to metanetworks list
  metanetwork$networkGroup <- bound(metanetwork)

metanetwork$networkGroup
```

<br/>

### Nodes coordinates

Now we need to get the coordinates of the edges so that we can position them
on the graph. We also need to get coordinates for each network so that we can
build an internal, hierarchical structure to the distribution of edges on the
graph that can be used to plot links between edges. These should be distributed
as the user wish within the circle using different radius values. As with the
groups, we can add a gap between the edges at the beginning and end of the group
for visual differentiation between the group.

```{r nodePos, fig.width = 5, fig.height = 5, fig.align = 'center'}
  nodePos <- function(metanetwork, edgeRad = 0.975, groupRad = 0.5, gapEdge = 0.1, addGap = T) {
    # Add x and y columns to nodes and networkGroup data
      metanetwork$nodes$y <- metanetwork$nodes$x <- 0
      metanetwork$networkGroup$y <- metanetwork$networkGroup$x <- 0

    # Get coordinates for all networks
      for(i in 1:nrow(metanetwork$networkGroup)) {
        # Distribute points within each network space
          edgeDeg <- seq((metanetwork$networkGroup$lower[i] + (gapEdge/2)),
                         (metanetwork$networkGroup$upper[i] - (gapEdge/2)),
                         length = metanetwork$networkGroup$Freq[i])

        # Get position for each edge   
          nodePos <- coordCircle(theta = edgeDeg, radius = edgeRad)

        # Add to nodes data
          metanetwork$nodes$x[metanetwork$nodes$network == metanetwork$networkGroup$Var1[i]] <- nodePos$x
          metanetwork$nodes$y[metanetwork$nodes$network == metanetwork$networkGroup$Var1[i]] <- nodePos$y

        # Distribute network groups in space
          groupDeg <- mean(c(metanetwork$networkGroup$lower[i],metanetwork$networkGroup$upper[i]))

        # Get position for each group
          groupPos <- coordCircle(theta = groupDeg, radius = groupRad)

        # Add to group data
          metanetwork$networkGroup$x[i] <- groupPos$x
          metanetwork$networkGroup$y[i] <- groupPos$y
      }

    metanetwork
  }

  metanetwork <- nodePos(metanetwork, edgeRad = .875, groupRad = .5)

  head(metanetwork$nodes)
  head(metanetwork$networkGroup)

  # Visualize the positions of the nodes
  # Plot
    par(mar = c(0,0,0,0))
    plot0()    
    points(metanetwork$nodes$x, metanetwork$nodes$y, pch = 20, cex = 2)
    points(metanetwork$networkGroup$x, metanetwork$networkGroup$y, pch = 20, cex = 2)
```


## Visualization functions

Now we need a few functions to automate the visualization of a few components of
the circular plot.

### Plot groups

I actually wish to have a box around the edge of my graph identifying each
individual network and their name. To do this, we can use the `polygon`
and `plotrix::arctext` function.

```{r plotGroups, fig.width = 10, fig.height = 5, fig.align = 'center'}
# We can now use both functions to generate a box for each group on the plot
boxGroup <- function(metanetwork, rad1 = .95, rad2 = 1, colBox = NULL, names = NULL, colNames = NULL, addNames = T, ...) {
  # metanetwork = data list composed of 'nodes', 'links' & 'networkGroup'
  # rad1 = lower boundary for polygons
  # rad2 = upper boundary for polygons
  # colBox = color of boxes
  # names = names of individual networks
  # colNames = color of names
  # addNames = logical, add names of networks to graph
  if (!is.null(colNames) & length(colNames) == 1) {
    colNames <- rep(colNames, nrow(metanetwork$links))
  }

  if (!is.null(colBox) & length(colBox) == 1) {
    colBox <- rep(colBox, nrow(metanetwork$links))
  }

  for(i in 1:nrow(metanetwork$networkGroup)) {
    a <- coordCircle(theta = seq(metanetwork$networkGroup$lower[i],
                             metanetwork$networkGroup$upper[i],
                             length = 200),
                     radius = rad1)

    b <- coordCircle(theta = seq(metanetwork$networkGroup$upper[i],
                             metanetwork$networkGroup$lower[i],
                             length = 200),
                     radius = rad2)

    polygon(rbind(a, b, a[1L,]), col = colBox[i], ...)

    if (addNames) {
      middle <- mean(c(metanetwork$networkGroup$lower[i],
                       metanetwork$networkGroup$upper[i]))
      clockwise <- if (middle > pi) F else T
      plotrix::arctext(x = as.character(metanetwork$networkGroup$Var1[i]),
                       radius = mean(c(rad1,rad2)),
                       middle = middle,
                       col = colNames[i],
                       clockwise = clockwise,
                       font = 2)    
    }
  }
}

# Now we can plot
  par(mfrow = c(1,2), mar = c(0,0,0,0))
  plot0()
  boxGroup(metanetwork, addNames = F)
  plot0()
  boxGroup(metanetwork, rad1 = .5)
```

### Plot connections

Now we can plot the links between the nodes in our graph using the underlying
hierarchical structure identified using the `nodePos` function created earlier.
We can do this using the `xspline` function.

```{r plotLinks, fig.width = 7, fig.height = 7, fig.align = 'center'}
  plotLinks <- function(metanetwork, cols = NULL, ...) {
    if (!is.null(cols) & length(cols) == 1) {
      cols <- rep(cols, nrow(metanetwork$links))
    }

    for(i in 1:nrow(metanetwork$links)) {
      link <- metanetwork$links[i,]
      edgeFromID <- which(metanetwork$nodes$name == link$from)
      edgeToID <- which(metanetwork$nodes$name == link$to)
      groupFromID <- which(metanetwork$networkGroup$Var1 == metanetwork$nodes$network[edgeFromID])
      groupToID <- which(metanetwork$networkGroup$Var1 == metanetwork$nodes$network[edgeToID])

      if (metanetwork$nodes$network[edgeFromID] != metanetwork$nodes$network[edgeToID]) {
        linkPath <- rbind(metanetwork$nodes[edgeFromID, c('x','y')],
                          metanetwork$networkGroup[groupFromID, c('x','y')],
                          metanetwork$networkGroup[groupToID, c('x','y')],
                          metanetwork$nodes[edgeToID, c('x','y')])
      } else {
        linkPath <- rbind(metanetwork$nodes[edgeFromID, c('x','y')],
                          metanetwork$networkGroup[groupFromID, c('x','y')],
                          metanetwork$nodes[edgeToID, c('x','y')])
      }

      lines(xspline(linkPath$x, linkPath$y, shape = 1, draw=FALSE), col = cols[i], ...)
    }
  }

  par(mar = c(0,0,0,0))
  plot0()
  plotLinks(metanetwork, col = 'black')
  points(metanetwork$nodes$x, metanetwork$nodes$y)
```

</br>

## Plot metanetwork

```{r metanetwork, fig.width = 6, fig.height = 6, fig.align = 'center'}
# The plot
  par(mar = c(0,0,0,0))
  plot0()
  boxGroup(metanetwork)
  plotLinks(metanetwork, col = 'black')
  points(metanetwork$nodes$x, metanetwork$nodes$y, pch = 20, cex = 2)
```

<br/>


### Customize graph functions

Now we can build functions to customize the graph a little bit, *e.g.* add colors
to links and nodes and change the size of the nodes.

First, let's give a color to individual networks.

```{r colGroups, fig.width = 7, fig.height = 7, fig.align = 'center'}
colGroups <- function(metanetwork, colPal = gpu$insileco) {
  # Group colors
    metanetwork$networkGroup$cols <- colPal[1:nrow(metanetwork$networkGroup)]

  # Node colors
    metanetwork$nodes$cols <- NA
    for(i in 1:nrow(metanetwork$networkGroup)) {
      metanetwork$nodes$cols[metanetwork$nodes$network == metanetwork$networkGroup$Var1[i]] <- metanetwork$networkGroup$cols[i]
    }

  metanetwork
}

metanetwork <- colGroups(metanetwork, colPal = gpu$insileco)

head(metanetwork$nodes)
head(metanetwork$networkGroup)
```

</br>

Second, a function to give a size to nodes based on link frequency.

<br/>

```{r nodeSize}
# Add size to nodes based on frequency of links to each node
nodeSize <- function(metanetwork, freq = T) {  
    if (isTRUE(freq)) {
      nLink <- as.data.frame(table(c(metanetwork$links$from, metanetwork$links$to)), stringsAsFactors = F)
      colnames(nLink)[1L] <- 'name'
      metanetwork$nodes <- dplyr::left_join(metanetwork$nodes, nLink, by = 'name')
      metanetwork$nodes$cex <- (metanetwork$nodes$Freq / max(metanetwork$nodes$Freq))  
    } else {
      metanetwork$nodes$cex <- .33
    }

    return(metanetwork)
}

metanetwork <- nodeSize(metanetwork)
head(metanetwork$nodes)
```

<br/>

Finally, we can give a color to links or focus on certain individual networks in
the metanetwork, or focus on all the links for a single network.

<br/>

```{r linkFocus}
linkCol <- function(metanetwork, type = 'all', focus = NULL, colLinks = '#876b40', colShadow = '#f4f4f4') {
  # metanetwork = list composed of 'nodes', 'links' and 'networkGroup'
  # type        = type of colors:
  #                 'all' = all links with single color = `colLinks`
  #                 'focus' = focus on the links of identified network
  # focus       = character, name of network(s) to focus on;
  #                 if length(focus) == 1, all links towards a single network
  #                 if length(focus) > 1, links focused on identified networks
  # colLinks    = color of links of `type` == 'all'
  # colShadow   = color of links that we are not focused on

  # Function
  if (type == 'all') {
    metanetwork$links$cols <- colLinks
  }

  if (type == 'focus' & length(focus) == 1) {
    # Box colors
    focusID <- metanetwork$networkGroup$Var1 %in% focus
    colBox <- metanetwork$networkGroup$cols
    metanetwork$networkGroup$cols[!focusID] <- colShadow
    metanetwork$networkGroup$colNames <- colBox
    metanetwork$networkGroup$colNames[focusID] <- colShadow

    # Link colors
    # metanetwork$links$cols <- paste0(colShadow, 88)
    metanetwork$links$cols <- colShadow
    linkCol <- data.frame(from = metanetwork$nodes$network[match(metanetwork$links$from,
                                                                 metanetwork$nodes$name)],
                          to = metanetwork$nodes$network[match(metanetwork$links$to,
                                                                metanetwork$nodes$name)],
                          stringsAsFactors = F)

    linkID <- linkCol$from %in% focus & linkCol$to %in% focus
    metanetwork$links$cols[linkID] <- metanetwork$networkGroup$cols[focusID] # "cannibalism"

    linkID <- (linkCol$from %in% focus | linkCol$to %in% focus) & !linkID
    cols <- paste0(linkCol$from[linkID], linkCol$to[linkID])
    cols <- gsub(focus, '', cols)
    cols <- match(cols, metanetwork$networkGroup$Var1)
    cols <- metanetwork$networkGroup$colNames[cols]
    metanetwork$links$cols[linkID] <- cols
  }

  if (type == 'focus' & length(focus) > 1) {
    # Box colors
    focusID <- metanetwork$networkGroup$Var1 %in% focus
    colBox <- metanetwork$networkGroup$cols
    metanetwork$networkGroup$cols[!focusID] <- colShadow
    metanetwork$networkGroup$colNames <- colBox
    metanetwork$networkGroup$colNames[focusID] <- colShadow

    # Link colors
    metanetwork$links$cols <- colShadow
    linkCol <- data.frame(from = metanetwork$nodes$network[match(metanetwork$links$from,
                                                                 metanetwork$nodes$name)],
                          to = metanetwork$nodes$network[match(metanetwork$links$to,
                                                                metanetwork$nodes$name)],
                          stringsAsFactors = F)

    linkID <- linkCol$from %in% focus & linkCol$to %in% focus
    metanetwork$links$cols[linkID] <- colLinks
  }

  # Add transparency
  metanetwork$links$cols <- paste0(metanetwork$links$cols, '66')

  metanetwork
}

metanetwork <- linkCol(metanetwork, type = 'focus', focus = c('Species','Drivers'))
head(metanetwork$links)
head(metanetwork$networkGroup)
```

<br/>

## Metanetwork function

Now we could wrap all of this in a single function.

```{r function, fig.width = 10, fig.height = 10, fig.align = 'center'}
plotMetanetwork <- function(metanetwork,
                            rad1 = .925,
                            rad2 = 1,
                            sizeEdge = T,
                            colPal = gpu$insileco,
                            type = 'all',
                            focus = NULL,
                            colLinks = '#876b40',
                            colShadow = '#f4f4f4',
                            shadowEdge = T
                          ) {

    # Metanetwork = list composed of 'nodes' and 'links'
    # rad1 = lower boundary for individual networks
    # rad2 = upper boundary for individual networks
    # colPal = color palette
    # colLinks = color for links

    # Function
    # Boundaries of individual networks
      metanetwork$networkGroup <- bound(metanetwork)

    # Node coordinates
      metanetwork <- nodePos(metanetwork, edgeRad = .875, groupRad = .5)

    # Colors
      metanetwork <- colGroups(metanetwork, colPal = colPal)

    # Node size
      metanetwork <- nodeSize(metanetwork, freq = sizeEdge)

    # Link col
      metanetwork <- linkCol(metanetwork, type = type, focus = focus, colLinks = colLinks, colShadow = colShadow)

    # Plot
    par(mar = c(0,0,0,0))
    plot0()
    boxGroup(metanetwork,
             rad1 = rad1,
             colBox = metanetwork$networkGroup$cols,
             colNames = metanetwork$networkGroup$colNames,
             border = 'transparent')
    plotLinks(metanetwork,
              col = metanetwork$links$cols)
    if (shadowEdge) {
      points(metanetwork$nodes$x,
             metanetwork$nodes$y,
             pch = 20,
             cex = (metanetwork$nodes$cex * 5),
             col = '#d7d7d7')  
    }

    points(metanetwork$nodes$x,
           metanetwork$nodes$y,
           pch = 20,
           cex = (metanetwork$nodes$cex * 3),
           col = metanetwork$nodes$cols)
}

# The data
metanetwork <- vector('list', 0)
metanetwork$nodes <- nodes
metanetwork$links <- links

# The plot
par(mfrow = c(2,2))
plotMetanetwork(metanetwork)
plotMetanetwork(metanetwork, type = 'focus', focus = 'Species')
plotMetanetwork(metanetwork, type = 'focus', focus = c('Species', 'Drivers'))
plotMetanetwork(metanetwork, type = 'focus', focus = c('Species', 'Drivers', 'Managers'))
opt <- options()
```
