---
title: Trick or Tips 003 {R}
author: [kevin, david]
date: 2018-02-11
tags: [R, tips, trickortips]
rpkgs: [base, utils, graphics]
tweet: Third post of the 'Trick or Tips' series
draft: false
output:
  blogdown::html_page:
    dev: svg
---



`r htmltools::HTML("{{< trickortips >}}")`


```{r codeChunkSetUp, echo=FALSE}
source('../../static/Rscript/codeChunkSetUp.R')
```


<br>

## The `apropos()` function

A powerful way to look for a function you can barely remember the name of
directly in R, *i.e* without googling!

```{r apropos1}
apropos('Sys')
```

You can also take advantage of regular expressions to narrow down you research:

```{r apropos2}
apropos('^Sys')
```

Or even better:

```{r apropos3}
apropos('^Sys.*time$', ignore.case = FALSE)
```


<br>

## The `table()` function

Oftentimes we wish to extract the frequency of certain elements in a dataset.
There is a very useful function that allows us to achieve this quite efficiently:
`table()`. Let's see how this works:

```{r table1}
df <- data.frame(data = sample(1:5, 20, replace = T))
table(df$data)
```

You can also get the frequency for a data.frame with multiple columns. For example,
if you observed species at a site throughout multiple years and wanted to know
the frequency of observations per species per year:


```{r table2}
df <- data.frame(observations = paste0('species', sample(1:5, 50, replace = T)),
                 year = sort(sample(2015:2018, 50, replace = T)))
table(df)
```

You can actually do so for more than two columns.

```{r table3}
df$atr1 <- rep(c("val1", "val2"), each = 25)
tb <- table(df)
tb
```

As you can see, in such case, you will have to deal with arrays:

```{r table4}
tb[, , 1]
```

With further development and by combining `table()` with `paste0()` (see
[fish and tips 001](https://insileco.github.io/2017/06/30/trick-or-tips-001-r/)
for an explanation of this useful function!), you can create your desired data.frame:

```{r table5}
as.data.frame(table(paste0(df$year,'_',df$observations)))
```


<br>

## Everything but 0

This is a well-known trick for developers that may be useful for many beginners.
In R when performing a logical test, every numeric is considered as `TRUE` but
`0` (which is `FALSE`):

```{r test1, results = "hold"}
0 == FALSE
!0
!1
!7.45
```

This can actually be very helpful, for instance when we are testing whether
or not a vector is empty!

```{r test2, results = "hold"}
vec0 <- 1:7
vec1 <- vec0[vec0>5]
vec2 <- vec0[vec0>7]
!(length(vec1))
!(length(vec2))
```




<br>

## `expand.grid()` vs.`combn()`

If you often create empty data.frame, you are very likely already familiar
with the `expand.grid()` function:


```{r expand1}
expand.grid(LETTERS[1:4], LETTERS[5:6])
```

But if you are looking for unique combinations (think about all combinations of
games in a tournament of four team), you may feel that `expand.grid()` is not
what you need:

```{r expand2}
expand.grid(LETTERS[1:4], LETTERS[1:4])
```

In comes `combn`:

```{r expand3}
combn(LETTERS[1:5], 2)
```

As you can see you need to specify the number of elements in the combination as
`combn` can compute all combination

```{r expand4}
combn(LETTERS[1:5], 4)
```

Also if you want a dataframe, a small extra step is required:

```{r expand5}
as.data.frame(t(combn(LETTERS[1:5], 2)))
```


<br>

## Writing outside the margins

If you are always thinking outside the box you may want to learn how to plot
something outside the margins! This is possible using the `xpd` parameter of
the `par()` function.


```{r xpd2}
par(mfrow = c(1, 2))
plot(c(0, 2), c(0, 2))
lines(c(-1,3), c(1,1), lwd = 4)
##
par(xpd = TRUE)
plot(c(0, 2), c(0, 2))
lines(c(-1,3), c(1,1), lwd = 4)
```

<br>

### See you next post post!

<br>
