---
title: R in Space - Import spatial objects
author: [david, marieh, nicolas, kevin, elliot, steve]
reviewer: []
date: 2018-04-09
tags: [R in Space, R, Spatial]
rpkgs: [sf, raster]
draft: true
navposts:
  prev: rinspace_spobjinr
  home: rinspace_homepage
  next: rinspace_basicmapping
estime: 16
---


## Reading and writing spatial objects with `sf` and `raster`

Spatial data are encoded in various [GIS file formats](https://en.wikipedia.org/wiki/GIS_file_formats)
such as Shapefiles (.shp), Geodatabase (.gdb), GeoPackage (.gpkg) and GeoJSON (.geojson),
GeoTIFF (.tiff). In this post we show how to read commonly used formats and
import the data they include in you R session. We start by describing the steps
to turn a set of coordinates included in a `.csv` file into a spatial object and
write it to a GIS file format. We then exemplify the use of two functions:
`st_read()` from the `sf` package and `raster()` from the `raster` package and
import different GIS file available on the [Portail de données ouvertes de
Montreal](http://donnees.ville.montreal.qc.ca/dataset).

Let us start by loading the packages required:

```{r packages, message=F}
library(sf)
library(raster)
```

## Import spatial data from .csv file

Spatial data are sometimes stored in a text file format (.txt or .csv), in such
case columns providing the coordinates must be included and the user must know
the CRS (Coordinate Reference System). A `sf` object allows us to store both
the coordinates of each point as well as associated attribute data, *i.e.*
columns describing each feature in the spatial object.

The dataset we work with below represents sampling points of a monitoring
program of water quality in Montreal (available [here](http://donnees.ville.montreal.qc.ca/dataset/rsma-points-d-echantillonnage-ruisso/resource/adad6c48-fb48-40fc-a031-1ac870d978b4)).


```{r read_points}
# Create a new directory to download data
dir.create("data")

# Download csv file from web page in your working directory
if (!file.exists("data/ruisso.csv"))
  download.file("http://donnees.ville.montreal.qc.ca/dataset/86843d31-4251-4002-b10d-620aaa751092/resource/adad6c48-fb48-40fc-a031-1ac870d978b4/download/scri03.-infor03.02-informatique03.02.07-donneesouvertesrsmaruissostationsstations_ruisso.csv", destfile = "data/ruisso.csv")

# Read csv file in R
pts <- read.csv("data/ruisso.csv", header = T, dec = ",")
```



### Convert a data frame to simple feature objects

The data frame containing sampling points can be converted into simple feature
objects using `st_as_sf()`. To do so, we need to specify where coordinates are,
*i.e.* columns `LATITUDE` and `LONGITUDE` as well as the CRS. In our example,
the CRS is specified in the [metadata of the data set](http://donnees.ville.montreal.qc.ca/dataset/rsma-points-d-echantillonnage-ruisso/resource/adad6c48-fb48-40fc-a031-1ac870d978b4): the [datum](http://www.nrcan.gc.ca/node/9791) is WGS84 and the [EPSG](http://www.epsg.org/) is **4326**.
Remember that without the CRS, you cannot locate correctly your coordinates on the
Earth’s surface. In the package `sf`, the reference system can be define using
the `proj4` format or directly with the EPSG code.


```{r convert_to_sf, eval=T, message = F}
pts_sf <- st_as_sf(x = pts,
  coords = c("LONGITUDE", "LATITUDE"),
  crs = 4326)
head(pts_sf)
```

As you can see, we now have a `MULTIPOINT` geometry, and the spatial information
is now stored in a simple feature list-column (`sfc`). The other columns contain
all the attributes related to the sample points.

The default plot of a simple features object is a multi-facet of all attributes.

```{r plot_pts, eval=T}
plot(pts_sf)  
```

As you can see, instead of creating a single map, as with sp object, the
default plot of sf object creates multiple maps, one for each attribute,
which can sometimes be useful for exploring the spatial distribution of different
variables.


### Export your points in a Shapefile

We can write a simple features object to a file (*e.g.* a shapefile) using the
`st_write()` function in `sf`, which needs at least two arguments, the object
and a filename:

```{r export, warnings = F}
st_write(pts_sf, "data/pts_sf.shp", delete_dsn = T)
```

Note that `st_write()` selected one driver base on the file extension provided.
The driver can also be made explicit using the `driver` argument like so:
`st_write(pts_sf, "data/pts_sf.shp", driver = "ESRI Shapefile")`. Last but not
least, in order to have details about drivers available, have a look at the data
frame `st_drivers()` returns.



## Import Shapefile data

The shapefile contains polygons delimiting the woods of the Montreal
agglomeration and information about the forest composition (found
[here](http://donnees.ville.montreal.qc.ca/dataset/29791562-f050-401e-b405-5c1fbf427f65)).


```{r read_poly}
# Download shapefile from web page in your working directory
if (!file.exists("data/bois.zip"))
  download.file("http://donnees.ville.montreal.qc.ca/dataset/29791562-f050-401e-b405-5c1fbf427f65/resource/9fa20d3a-5dee-43d6-9bc9-5d86fe225e16/download/bois.zip", destfile = "data/bois.zip")

unzip("data/bois.zip", exdir = "data")

# Read shapefile in R
bois <- st_read(dsn = "data/", layer = "bois")

head(bois)
```

The `bois` dataset has been turned into a `MULTIPOLYGON` object and has the same
CRS (EPSG: 4326) than the sample points we have manipulated above. This allows
us to work directly with the two objects otherwise we should have transformed
one dataset using the CRS of the other. To plot only the geometry and not all
attributes, we retrieve the geometry list-column using `st_geometry()`:

```{r plot_poly_geom}
plot(st_geometry(bois))
```

To plot the polygons with a thematic color scale according to one attribute of
interest, we actually subset the object (here we use the name of the column):

```{r plot_poly_area}
plot(bois["Shape_Area"])
```


## Import GeoJSON file

This GeoJSON dataset contains watercourses (stream, river) and main ditches of
the Montreal agglomeration (available [here](http://donnees.ville.montreal.qc.ca/dataset/c128aff5-325c-4599-ab66-1c9d0b3abc94)). Hence, it is a `MULTILINE` object.

```{r read_geojson}
# Download shapefile from web page in your working directory
if (!file.exists("data/courseau.geojson"))
  download.file("http://donnees.ville.montreal.qc.ca/dataset/c128aff5-325c-4599-ab66-1c9d0b3abc94/resource/0f64976e-19c1-4d29-bcc5-4b663a392617/download/courseau.geojson", destfile="data/courseau.geojson")

# For GeoJSON, dsn may be the character string holding the geojson data

courseau <- st_read(dsn = "data/courseau.geojson")

head(courseau)

plot(st_geometry(courseau))
```



## Import raster data

We now import raster data use a [.tif file](https://en.wikipedia.org/wiki/TIFF)
file of a canopy index from Montreal (found [here](http://cmm.qc.ca/donnees-et-territoire/observatoire-grand-montreal/produits-cartographiques/donnees-georeferencees/)).
The canopy index, computed by the Montreal Metropolitain Community (CMM) from
an NDVI index and an elevation surface model, represents the low vegetation
cover, the high vegetation cover of more than 3 meters (the canopy), the low
mineral surfaces and the high mineral surfaces more than 3 meters (roof).


```{r read_tif, warning=F}
# Download tif file from web page in your working directory
if (!file.exists("data/canopee.zip"))
  download.file("http://cmm.qc.ca/fileadmin/user_upload/geomatique/IndiceCanopee/2015/660_Canopee2015_3m.zip", destfile="data/canopee.zip")

unzip("data/canopee.zip", exdir = "data")

# Read tif in R using raster
# The file named "660_CLASS_3m.tif" contains the canopy index for all the Montreal area, so we can read this file only

canopee_mtl <- raster("data/660_CLASS_3m.tif")
```

The canopy index raster has values from 1 to 5, has `nrow(canopee_mtl)` pixels
by row and `ncol(canopee_mtl)` pixels by column. Note that raster uses the
`proj4string` representation of the coordinate reference system.

```{r info_tif, warning=F}
canopee_mtl
```

Similar to the sf package, raster also provides `plot` methods for its own classes.

```{r plot_tif}
plot(canopee_mtl)
```
