---
title: "R in Space - About spatial data"
author: [david, marieh, nicolas, kevin, elliot, steve]
date: 2018-04-12
tags: [R in Space, R, Spatial]
rpkgs: [raster, sf, sp]
tweet: "R in Space - About spatial data"
draft: false
navposts:
  prev: rinspace_refs
  home: rinspace_homepage
  next: rinspace_import
output:
  rmarkdown::html_page:
    toc: true
    fig_width: 3
    dev: svg
---


```{r codeChunkSetUp, echo=FALSE}
source("../../../../static/Rscript/codeChunkSetUp.R")
```


<br><br>

Spatial data refer to phenomenon or information that can be observed geographically.
What a definition! In short spatial data concern every information that could
be reported on a map, *i.e.* on a two-dimensions geographical coordinates system.
Two kinds of spatial data exist: **vector data** and **raster data**. Generally
vectors refer to discrete locations, *i.e.* objects with boundaries (for
instance a city, a road, a country) whereas raster data refer to continuous
phenomenon that can be observed everywhere, but without natural boundaries (for
instance the surface temperature). Let's take a look at their characteristics.


## Vector data

Vector data consist of the description of a spatially explicit phenomenon
geometry (position and coordinates of its boundaries in a specific spatial
referential). In addition to this geometry, vector data may contain variables
(or attributes) with additional information about the phenomenon at each
location. For instance, cities of a country are vector data (points) and
attributes could be cities names, population sizes, etc. Borders of a country
are also vector data (polygons) and could contain the name of the country,
the population for a given year, and the mean number of crimes as attributes.

There are main symbol types for vector data: **points**, **lines** and **polygons**.
The geometry of these data structures consists of sets of geographical
coordinate pairs (longitude, latitude) in a specific Coordinate Reference System
(CRS).


### Points data

Points are the simplest vector object. Each point has one two-dimensional
coordinates, and several associated variables. For instance, a point might
represent a location where an animal was trapped, and the attributes could
include the capture date, the size, the sex, and information about the physical
environment.

<br />

```{r vector-data, echo=FALSE, eval=TRUE, fig.width=8.6, fig.height=2.86, dpi=300, out.width=810, out.height=270, fig.align='center'}
library(raster)

corse  <- readRDS("data/corse-admin.rds")
rivers <- readRDS("data/corse-rivers.rds")
cities <- readRDS("data/corse-cities.rds")

par(
  mar      = rep(1.5, 4),
  xaxs     = "i",
  yaxs     = "i",
  mfrow    = c(1, 3),
  col.axis = "#555555",
  family   = "serif"
)

# Vector - POINTS
plot(corse, col = NA, border = NA)
points(cities[ , c("lon", "lat")], pch = 21, bg = par()$col.axis, col = "white", cex = 2)
text(8, 42.8, "Vector data\n(POINTS)", font = 2, col = par()$col.axis, pos = 4, cex = 1.25)

par(mgp = c(3, .5, 0))
axis(1, lwd = 0, lwd.ticks = 1, col = par()$col.axis)
par(mgp = c(3, .695, 0))
axis(4, label = FALSE, lwd = 0, lwd.ticks = 1, col = par()$col.axis, las = 1)
mtext(side = 2, line = .15, "Latitude (°North)", font = 1, col = par()$col.axis, cex = .75)
mtext(side = 3, line = .15, "Longitude (°East)", font = 1, col = par()$col.axis, cex = .75)
box(col = par()$col.axis)

# Vector - LINES
plot(corse, col = NA, border = NA)
plot(rivers, add = TRUE, col = "#96DAE8", lwd = 3)
plot(rivers, add = TRUE, col = "#009BC9", lwd = 1)
plot(corse, col = NA, border = "white", add=T, lwd = 3)
text(8, 42.8, "Vector data\n(LINES)", font = 2, col = par()$col.axis, pos = 4, cex = 1.25)

par(mgp = c(3, .5, 0))
axis(1, lwd = 0, lwd.ticks = 1, col = par()$col.axis)
par(mgp = c(3, .695, 0))
axis(2, lwd = 0, lwd.ticks = 1, col = par()$col.axis, las = 1)
axis(4, label = FALSE, lwd = 0, lwd.ticks = 1, col = par()$col.axis, las = 1)
mtext(side = 3, line = .15, "Longitude (°East)", font = 1, col = par()$col.axis, cex = .75)
box(col = par()$col.axis)

# Vector - POLYGONS
plot(corse, col = "#aaaaaa", border = NA)
text(8, 42.8, "Vector data\n(POLYGONS)", font = 2, col = par()$col.axis, pos = 4, cex = 1.25)

par(mgp = c(3, .5, 0))
axis(1, lwd = 0, lwd.ticks = 1, col = par()$col.axis)
par(mgp = c(3, .695, 0))
axis(2, lwd = 0, lwd.ticks = 1, col = par()$col.axis, las = 1)
mtext(side = 4, line = .15, "Latitude (°North)", font = 1, col = par()$col.axis, cex = .75)
mtext(side = 3, line = .15, "Longitude (°East)", font = 1, col = par()$col.axis, cex = .75)
box(col = par()$col.axis)
```



<br />

### Lines data

Lines are the second type of vector data. A line is a shape consisting of one
or several segments (or paths) and each segment has two noteworthy points
(or vertex): a starting point and an ending point. Note that the ending point of
a segment will have the same coordinates as the starting point of the following
segment in the case of a line consisting of connected segments. In other words,
a line with one segment is defined by two coordinates pairs (longitude and
latitude). A line with two connected segments is defined by three coordinates
pairs and a line with two isolated segments is defined by four coordinates pairs.
Thus, the representation of a line is very similar to that of a collection of
points. The main difference is that the ordering of the points is important,
because we need to know which points should be connected two-by-two.


### Polygons data

Polygons are closed spatial lines where the coordinates of the starting point of
the first segment is identical to the ending point of the last segment. The
geometry of a polygon is very similar to that of a line but polygons have three
characteristics:

1. a polygon cannot self-intersect (whereas a line can),

2. a polygon can contain hole (think about the state of Vatican overlapping the
  country of Italy but considered as a separated polygon),

3. as a closed feature, a polygon has an inside and a border.

<br />

Vector data are generally stored in **ESRI Shapefile**, **GeoJSON**, **KML**
or **Simple Features** files. In R, two main packages exist to handle spatial
vector data: `sp` and `sf`. Note that the package `rgdal` will be used to
import/export vector data (go next post to learn more).


## Raster data

Raster data are commonly used to represent _continuous_ phenomenon that can be
observed everywhere, but without natural boundaries (but with artificial
boundaries defined by the user). A raster consists of a grid of equally sized
cells (or pixels) that all have a values (or a missing value) for one single
variable.

Unlike vector data, the geometry of raster data is not explicitly stored as
coordinates. Rather it is implicitly set by setting the spatial extent and the
number of rows and columns of a regular grid. From this spatial information,
the size of the raster cells (spatial resolution) can be computed. Working with
raster data will therefore be more efficient than working with polygons data.

Raster can be used to represent a very broad range of data: continuous
(temperature values), discrete (habitat classes) or even binary (species occurrence)
variables.


```{r raster-data, echo=FALSE, eval=TRUE, fig.width=8.6, fig.height=2.86, dpi=300, out.width=810, out.height=270, fig.align='center'}
library(rgdal)

corse      <- readRDS("data/corse-admin.rds")
elevation  <- readRDS("data/corse-elevation.rds")
landcover  <- readRDS("data/corse-landcover.rds")
occurrence <- readRDS("data/corse-occurrence.rds")

par(
  mar      = rep(1.5, 4),
  xaxs     = "i",
  yaxs     = "i",
  mfrow    = c(1, 3),
  col.axis = "#555555",
  family   = "serif"
)

# CONTINUOUS VARIABLE
plot(corse, col = NA, border = NA)
plot(elevation, add = TRUE, axes = FALSE, legend = FALSE, col = heat.colors(n = 255)[255:1])
# plot(corse, col = NA, border = "#aaaaaa", add = TRUE)
text(8, 42.85, "Continuous raster", font = 2, col = par()$col.axis, pos = 4, cex = 1.25)

par(mgp = c(3, .5, 0))
axis(1, lwd = 0, lwd.ticks = 1, col = par()$col.axis)
par(mgp = c(3, .695, 0))
axis(4, label = FALSE, lwd = 0, lwd.ticks = 1, col = par()$col.axis, las = 1)
mtext(side = 2, line = .15, "Latitude (°North)", font = 1, col = par()$col.axis, cex = .75)
mtext(side = 3, line = .15, "Longitude (°East)", font = 1, col = par()$col.axis, cex = .75)
box(col = par()$col.axis)

# DISCRETE VARIABLE
plot(corse, col = NA, border = NA)
plot(landcover, add = TRUE, axes = FALSE, legend = FALSE)
# plot(corse, col = NA, border = "#aaaaaa", add = TRUE)
text(8, 42.85, "Discrete raster", font = 2, col = par()$col.axis, pos = 4, cex = 1.25)

par(mgp = c(3, .5, 0))
axis(1, lwd = 0, lwd.ticks = 1, col = par()$col.axis)
par(mgp = c(3, .695, 0))
axis(2, lwd = 0, lwd.ticks = 1, col = par()$col.axis, las = 1)
axis(4, label = FALSE, lwd = 0, lwd.ticks = 1, col = par()$col.axis, las = 1)
mtext(side = 3, line = .15, "Longitude (°East)", font = 1, col = par()$col.axis, cex = .75)
box(col = par()$col.axis)

# BINARY VARIABLE
plot(corse, col = NA, border = NA)
plot(occurrence, add = TRUE, axes = FALSE, legend = FALSE, col = c("#ffb770", "#dc1023"))
# plot(corse, col = NA, border = "#aaaaaa", add = TRUE)
text(8, 42.85, "Binary raster", font = 2, col = par()$col.axis, pos = 4, cex = 1.25)

par(mgp = c(3, .5, 0))
axis(1, lwd = 0, lwd.ticks = 1, col = par()$col.axis)
par(mgp = c(3, .695, 0))
axis(2, lwd = 0, lwd.ticks = 1, col = par()$col.axis, las = 1)
mtext(side = 4, line = .15, "Latitude (°North)", font = 1, col = par()$col.axis, cex = .75)
mtext(side = 3, line = .15, "Longitude (°East)", font = 1, col = par()$col.axis, cex = .75)
box(col = par()$col.axis)
```

<br />

Several file formats exist to store raster data. The most commons are: **GeoTIFF**, **NetCDF**, **grd** and **ascii** formats. Even the package `rgdal` can be used to import/export raster data, we will prefer the package `raster`.





## Vector objects in R

In this section we present two packges: `sp` and `sf`. The `sp` package actualy
defines classes for both vector and raster objects. Below, we
however focus on the vector ones and so we do not detail `SpatialGrid` and
`SpatialPixels` objects. Also note that `sf` "[...] aims at succeeding `sp` in
the long term" ([Simple Features for R](https://cran.r-project.org/web/packages/sf/vignettes/sf1.html),
`sf` vignette).

Let first create a data frame `mydata`:

```{r sim_points, results='asis'}
mylon <- -82+2*runif(20)
mylat <- 42+2*runif(20)
mydata <- data.frame(
  lon = mylon,
  lat = mylat,
  var1 = rnorm(20),
  var2 = 10*runif(20)
)
```

Let's have a look at thus data frame:

```{r show_points, results='asis', echo=F}
knitr::kable(head(mydata))
```






### Package `sp`

#### Classes

The table below includes a description of the classes for points, lines and polygons.
Basically all these classes work the same way. For instance, in order to define a
`SpatialPointsDataPoints` object, three elements are required: a set of coordinates,
a Coordinate Reference System (CRS) and an attribute table. Intermediate
class are also defined (for instance points + CRS = `SpatialPoints`) and
the name of the class is also the name of the function to be called.

Classes and functions       | Contents
--------------------------- | ---------------
Points                      | list of points (set of coordinates)
SpatialPoints               | list of points + CRS
SpatialPointsDataPoints     | list of points + CRS + attribute table
Line                        | a line (set of coordinates)
Lines                       | list of lines
SpatialLines                | list of lines + CRS
SpatialLinesDataFrame       | list of lines + CRS + attribute table
Polygon                     | a line (set of coordinates)
Polygons                    | list of lines
SpatialPolygons             | list of lines + CRS
SpatialPolygonsDataFrame    | list of lines + CRS + attribute table



#### `SpatialPointsDataFrame`

As an more tangible example, let's now create a `SpatialPointsDataFrame`
based on our data frame `mydata`.

```{r sppoint, message = F}
library(sp)
mysp <- SpatialPointsDataFrame(
  coords = mydata[, 1:2],
  data = mydata[, 3:4],
  proj4string = CRS(
    "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0"
  )
)
```

Note that [proj4](https://github.com/OSGeo/proj.4) is used and we therefore
wrote a string that describes the CRS and that proj4 understands. Below are
listed some properties of the object we have defined.



```{r sppoint2, results="hold"}
isS4(mysp)
class(mysp)
slotNames(mysp)
dim(mysp)
```

Basically, it is a [S4 object](http://adv-r.had.co.nz/S4.html) of class
`SpatialPointsDataFrame`. All slot names refer to attribute that are accessible
*via* and `@`:


```{r sppoint3}
mysp@proj4string
head(mysp@data)
```

In order to change projection, the user must call `spTransform()`, like so:

```{r spTransform}
(mysp2 <- spTransform(mysp, CRS=CRS("+proj=merc +ellps=GRS80")))
```


### Package `sf`

Below is a very short overview of classes in `sf`, the reader that requires
further explanation would find more details on the
[first vignette of sf](https://cran.r-project.org/web/packages/sf/vignettes/sf1.html#sf-objects-with-simple-features).
Basically three classes are defined: `sf`, `sfc` and `sfg`.


#### Class `sf`

```{r sf1_a}
library(sf)
pts_sf <- st_as_sf(
  x = mydata,
  coords = c("lon", "lat"),
  crs = 4326)
```

Let's examine its class

```{r sf1_b}
class(pts_sf)
```

#### Class `sfc`

```{r sf2}
pts_sfc <- st_geometry(pts_sf)
class(pts_sfc)
```


#### Class `sfg`

```{r sf3}
(x <- st_point(c(1,2)))
class(x)
```

#### How to import a `sp` object

`st_as_sf()` can also be used to convert a `sp` object into a `sf` one.

```{r sf4}
st_as_sf(mysp)
```





## Raster objects in R

The R package `raster` provides three main classes of raster object (more details
[here](http://www.rspatial.org/spatial/rst/4-rasterdata.html#rasterlayer)):

1. **RasterLayer** imports a single-layer (variable) raster,
2. **RasterStack** imports in one single object several single-layer (variable) rasters stored in one or different files,
3. **RasterBrick** imports in one single object several single-layer (variable) rasters stored in one single file.

Using **RasterStack** and **RasterBrick** requires that the geometry of all
raster data is equal.


Package `raster` define three classes of rater object we detail below.


### RasterLayer

```{r raster, message = F}
library(raster)
val1 <- matrix(runif(100*100,0,10), ncol = 100, nrow = 100)
ras1 <- raster(
    val1,
    crs = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"),
    xmn = -82, xmx = +80, ymn = +42, ymx = +44
  )
class(ras1)
dim(ras1)
head(values(ras1))
projection(ras1)
```



### RasterStack and RasterBrick

http://www.rspatial.org/spatial/rst/4-rasterdata.html#rasterstack-and-rasterbrick

Let's first create another raster (with the same CRS)

```{r raster3}
val2 <- matrix(rnorm(100*100), ncol = 100, nrow = 100)
ras2 <- raster(
    val2,
    crs = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"),
    xmn = -82, xmx = +80, ymn = +42, ymx = +44
  )
class(ras2)
```

Let's `stack()` `ras1` and `ras2`:

```{r raster4}
sta1 <- stack(ras1, ras2)
class(sta1)
```

Let's `brick()` them:

```{r raster5}
bri1 <- brick(ras1, ras2)
class(bri1)
```
